EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-08-21 jlamp/jlamp.tex
pruned e separati

 \smallskip 
 We can interpret an arbitrary hypergaph as a \emph{typed} or \emph{colored} monoidal signature: the nodes provide the \emph{types} or \emph{colors}, while hyperedges can be thought of as operations, with arity and coarity provided by the source and target functions, respectively. This suggests the following definition, which is used later on.
 % in Section \ref{par:tg}\label{sub:hyp_alg}.

@@

		Let $\mathcal{H}$ be $(\emptyset, 1, ?_{1^\star}, ?_{1^\star})$, which is clearly a separated hypergraph. Now,  $\shyp(\mathcal{H}, \sp(\mathcal{D}))$ has the same cardinality of  $\shyp(\mathcal{H}, \mathcal{G})$, which is $4$, while $\hyp(S(\mathcal{H}), \mathcal{D})$ is a singleton.
		\item Consider now the ``single loop'' $\mathcal{L}=(1, 1, \id{1}, \id{1})$. Then $\sp(\mathcal{L})$ is isomorphic to $(1, 2, \id{2}, \id{2})$ and so there are no arrows $\mathcal{L}\to S(\sp(\mathcal{L}))$. In particular, this shows that $\sp$ cannot be left adjoint to $S$. 
	\end{itemize} 
\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.27\linewidth}
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node[style=node] (v) at (0, 0){};
			\node[style=none] (vlab) at (0, 0.3){$0$};
			\node[style=small box] (times) at (-1, 1) {};
			\node[style=none] (timeslab) at (-1, 1.5) {$0$};
			\node[style=none] (timesin) at (-1, .7) {};
			\node[style=none] (timesout) at (-.7, 1){};
			\node[style=small box] (div) at (1, 1) {};
			\node[style=none] (divlab) at (1, 1.5) {$1$};
			\node[style=none] (divin) at (1, .7) {};
			\node[style=none] (divout) at (.7, 1){};

@@

		\begin{pgfonlayer}{edgelayer} 
			\draw[style=none,out=180, in= 270] (v) to (timesin.center);
			\draw[style=none,out=0, in= 270] (v) to (divin.center);
			\draw[style=none, in=135, out=345] (timesout.center) to (v);
			\draw[style=none, in=45, out=195] (divout.center) to (v);
			\draw[style=none, in=180, out=0] (a.center) to (zeroin.west);
			\draw[style=none, in=180, out=0] (c.center) to (unoin.west);
			\draw[style=none, in=180, out=0] (zeroout.east) to (b.west);
			\draw[style=none, in=180, out=0] (unoout.east) to (d.west);
		\end{pgfonlayer}
	\end{tikzpicture}
\end{minipage}

@@

@@

	Suppose that there exists $\mathcal{G}$ and arrows $(p_1, p_2)\colon \mathcal{Q}\to \mathcal{G}$ and $(q_1, q_2)\colon \mathcal{K}\to \mathcal{G}$ which make the square a pushout in $\shyp$. 
	By construction we have
	\[s_{\mathcal{G}}(q_1(e)) = q^\star_2(s_{\mathcal{K}}(e)) = q^\star_2(j^\star_2(v))=p^\star_2(i^\star_2(v)) = p^\star_2(t^\star_{\mathcal{Q}}(e))= t_{\mathcal{G}}(p_2(e))\]
	and so $\mathcal{G}$ cannot be in $\shyp$, yielding a contradiction.
\end{exa}
\end{minipage} \hfill \begin{minipage}[r]{.5\linewidth} \vspace{-2cm}
\xymatrix@R=6pt@C=30pt{\begin{tikzpicture}
		\node[style=node](v) at (0, 0){};
	\node[style=none] at(0, 0.3){$v$};
	\end{tikzpicture}\ar[d]_{(i_1, i_2)} \ar[r]^-{(j_1, j_2)}  &
	\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$j_2(v)$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$w$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, 0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw[style=none] (v) to (in.east); 
		\draw[style=none] (out.east) to (u.west);
	\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[d]^{(q_1, q_2)} \\\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$u$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$i_2(v)$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, -0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw[style=none] (v) to (in.east); 
			\draw[style=none] (out.east) to (u.west);
		\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[r]_-{(p_1, p_2)}&
	\mathcal{G}
	}
\end{minipage}

@@

\begin{thm} Let $\mathcal{G}$ be an hypergraph. Then %the following fact hold
%	\begin{enumerate}
%		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		%\item 
		$\lmo_{\mathcal{G}}$ is quasiadhesive.
%	\end{enumerate}
\end{thm}

@@

%\input{Hypergraphs_with_equivalences}
%\input{Labeled_hyp_with_eq}

@@

@@

	Now let $(h,k)$ be another morphism $\mathcal{H}\to \sp(\mathcal{G})$ such that $(\id{E_{\mathcal{G}}}, \epsilon_{G})\circ P(h, k)=(f,g)$. This entails immediately that $h =f$.  Let $v$ be a node $V_{\mathcal{H}}$. Since $\mathcal{H}$ is pruned, we have two cases
	\begin{itemize}
		\item $v$ occurs in $s_{\mathcal{H}}(e)$ for some $e\in E_{\mathcal{H}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{H}}}(s_{\mathcal{G}}(e))$ be such that $v=s_{\mathcal{G}}(e)_i$, then we have
		\[\hat{g}(v)=\hat{g}(s_{\mathcal{H}}(e)_i)=(g^\star(s_{\mathcal{H}}(e)))_i=s_{\sp(\mathcal{G})}(f(e))_i=(k^\star(s_{\mathcal{H}}(e)))_i=k(s_{\mathcal{H}}(e)_i)=k(v)\]
		\item $v$ occurs in $t_{\mathcal{H}}(e)$ for some $e\in E_{\mathcal{H}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{H}}}(t_{\mathcal{G}}(e))$ be such that $v=t_{\mathcal{G}}(e)_i$, then we have
		\[\hat{g}(v)=\hat{g}(t_{\mathcal{H}}(e)_i)=(g^\star(t_{\mathcal{H}}(e)))_i=t_{\sp(\mathcal{G})}(f(e))_i=(k^\star(t_{\mathcal{H}}(e)))_i=k(t_{\mathcal{H}}(e)_i)=k(v)\]
	\end{itemize}
In both cases $\hat{g}(v)=k(v)$ and so $\hat{g}= k$ as wanted.
\end{proof}

@@

@@

\section{Some properties of comma categories}
In this section we briefly recall the definition of the comma category \cite{mac2013categories} and some of its properties.
\begin{defi}\index{category!comma -}
	Let $L\colon \A\to \X$ and  $R\colon \B\rightarrow \X$ be two functors with the same codomain, the \emph{comma category} $\comma{L}{R}$ is the category in which

\noindent
\parbox{10.5cm}{\vspace{-.1cm}\begin{itemize}
		\item objects are triples $(A, B, f)$ with $A\in \A$, $B\in \B$, and $f\colon L(A)\rightarrow R(B)$; 
		\item a morphism $(A, B, f)\rightarrow (A', B', g)$ is a pair $(h, k)$ with $h\colon A\rightarrow A'$ in $\A$ and $k\colon B\rightarrow B'$ in $\B$ such that the diagram aside commutes.\end{itemize}}\hfill 
	\parbox{4cm}{\vspace{-.1cm}\xymatrix@C=25pt{L(A) \ar[r]^{L(h)} \ar[d]_{f} & L(A') \ar[d]^{g}\\ R(B) \ar[r]_{R(k)}& R(B')}}
\end{defi} 
We have two forgetful functors 	$U_L\colon \comma{L}{R}\to \A$ and $U_R\colon \comma{L}{R} \to \B$ given respectively by
\[
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{A}{h}{A'}
\end{split}\quad 
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{B}{k}{B'}
\end{split}
\]

Given $L\colon \A\to \X$ and $R\colon \B\to \X$, we can also consider their duals $L^{op}\colon \A^{op}\to \X^{op}$ and $R^{op}\colon \B^{op}\to \X^{op}$.  An arrow $f\colon L(A)\to R(B)$ in $\X$ is the same thing as an arrow $f\colon R^{op}(B)\to L^{op}(A)$ in $\X^{op}$. Thus $\comma{L}{R}$ and $\comma{R^{op}}{L^{op}}$ have the same objects. Moreover, the left square below commutes in $\X$ if and only if the right one commutes in $\X^{op}$
\[\xymatrix@C=25pt{L(A) \ar[r]^{L(h)} \ar[d]_{f} & L(A') \ar[d]^{g} & R(B') \ar[d]_{g} \ar[r]^{R(k)}& R(B) \ar[d]^{f}\\ R(B) \ar[r]_{R(k)}& R(B') & L(A') \ar[r]_{L(h)}  & L(A)}\]

@@

	\noindent \parbox{10cm}{
	\hspace{15pt}Thus there exists $f\colon L(A)\rightarrow R(B)$ fitting in the diagram on the right. Notice that $f$ is the unique arrow in $\X$ which makes $\left(a_D, b_D\right)$ an arrow $\left(A_D, B_D, f_D\right)\to \left(A, B, f\right)$ of $\comma{L}{R}$. If we show that $\left((A, B, f), \left\{(a_D, b_D)\right\}_{D\in \D}\right)$ is colimiting for $F$ we are done.}\hfill 
	\parbox{4cm}{\xymatrix@C=35pt@R=15pt{ L(A_D) \ar[r]^{L(a_D)} \ar[d]_{f_D}& L(A) \ar[d]^{f}\\ R(B_D) \ar[r]_{R(b_D)} &R(B)}}

@@

We can use \Cref{lim} to characterize monos in comma categories. 
\begin{cor}\label[corollary]{cor:mono}
	If $R$ preserves pullbacks then an arrow $(h,k)$ in $\comma{L}{R}$ is mono if and only if both $h$ and $k$ are monos.
\end{cor}
\begin{proof}
	$(\Rightarrow)$  If $(h,k)\colon (A,B,f)\to (A', B', g)$ is a mono then the first rectangle below is a pullback in $\comma{L}{R}$. By \Cref{lim} then also the other two squares are pullbacks, respectively in $\A$ and $\B$, proving that both $h$ and $k$ are monos
	\[\xymatrix{(A, B, f)  \ar[rr]^{\id{(A,B,f)}} \ar[d]_{\id{(A,B,f)}}&& (A, B, f)  \ar[d]^{(h,k)}& A \ar[r]^{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]^{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ (A, B, f)  \ar[rr]_{(h,k)}&& (A', B', g) &A \ar[r]_{h} & A' & B \ar[r]_{k} & B'}\]

%\newpage
	\smallskip\noindent 
	\parbox{4cm}{ \xymatrix{A \ar[r]^{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]^{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ A \ar[r]_{h} & A' & B \ar[r]_{k} & B'}}\hfill \parbox{8.5cm}{$(\Leftarrow)$ Since $h$ and $k$ are monos then we have the two pullback squares on the left. Thus by \Cref{lim} the pullback of $(h,k)$ along itself has isomorphisms as projections and so $(h,k)$ is mono. \qedhere}

@@

\begin{prop}\label[proposition]{prop:left}
	If $\A$ has initial objects and $L$ preserves them, then the forgetful functor $U_R\colon \comma{L}{R}\to \B$ has a left adjoint $\Delta$.
\end{prop}
\begin{proof} For an object $B\in \B$ we can define $\Delta(B)$ as $(0, B, ?_{R(B)})$, for $0$ is an initial object in $\A$ and $?_{R(B)}$ is the unique arrow $L(0)\to R(B)$. Let $\id{B}\colon B\to U_R(\Delta(B))$ be the identity,

>>>>

 \smallskip 
 We can interpret an arbitrary hypergaph as a \emph{typed} or \emph{colored} monoidal signature: the nodes provide the \emph{types} or \emph{colors}, while hyperedges can be thought of as operations, with arity and coarity provided by the source and target functions, respectively. This suggests the following definition, which is used later on\label{sub:hyp_alg}.
 % in Section \ref{par:tg}\label{sub:hyp_alg}.

@@

		Let $\mathcal{H}$ be $(\emptyset, 1, ?_{1^\star}, ?_{1^\star})$, which is clearly a separated hypergraph. Now,  $\shyp(\mathcal{H}, \sp(\mathcal{D}))$ has the same cardinality of  $\shyp(\mathcal{H}, \mathcal{G})$, which is $4$, while $\hyp(S(\mathcal{H}), \mathcal{D})$ is a singleton.
		\item Consider now the ``single loop'' $\mathcal{L}=(1, 1, \id{1}, \id{1})$. Then $\sp(\mathcal{L})$ is isomorphic to $(1, 2, \id{2}, \id{2})$ and so there are no arrows $\mathcal{L}\to S(\sp(\mathcal{L}))$. In particular, this shows that $\sp$ cannot be left adjoint to $S$. 
	\end{itemize} 
\end{rem}
\end{minipage}
\hfill
\begin{minipage}[r]{.27\linewidth}
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node[style=node] (v) at (0, 0){};
			\node[style=none] (vlab) at (0, 0.3){$0$};
			\node[style=small box] (times) at (-.8, 1) {};
			\node[style=none] (timeslab) at (-.8, 1.5) {$0$};
			\node[style=none] (timesin) at (-1.1, 1) {};
			\node[style=none] (timesout) at (-.5, 1){};
			\node[style=small box] (div) at (.8, 1) {};
			\node[style=none] (divlab) at (.8, 1.5) {$1$};
			\node[style=none] (divin) at (.5, 1) {};
			\node[style=none] (divout) at (1.1, 1){};

@@

		\begin{pgfonlayer}{edgelayer} 
			\draw[style=none,out=200, in= 180] (v) to (timesin.center);
			\draw[style=none,out=45, in= 200] (v) to (divin.center);
			\draw[style=none, in=135, out=345] (timesout.center) to (v);
			\draw[style=none, in=345, out=0] (divout.center) to (v);
			\draw[style=none, in=180, out=0] (a.center) to (zeroin.west);
			\draw[style=none, in=180, out=0] (c.center) to (unoin.west);
			\draw[style=none, in=180, out=0] (zeroout.east) to (b.west);
			\draw[style=none, in=180, out=0] (unoout.east) to (d.west);
		\end{pgfonlayer}
	\end{tikzpicture}
\end{minipage}

@@

\begin{rem}\label{rem:mprun}
	Let $(f,g)$ and $(h, k)$ be two morphisms $\mathcal{G}\rightrightarrows \mathcal{H}$ of $\phyp$ and suppose that $f=h$. We can use the absence of isolated nodes to show that $g=k$ too. Let $v\in V_{\mathcal{G}}$, we have two cases:
	\begin{itemize}
		\item $v$ occurs in $s_{\mathcal{G}}(e)$ for some $e\in E_{\mathcal{G}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ be such that $v=s_{\mathcal{G}}(e)_i$, then we have
		\[g(v)=g(s_{\mathcal{G}}(e)_i)=(g^\star(s_{\mathcal{G}}(e)))_i=s_{\mathcal{H}}(f(e))_i=s_{\mathcal{H}}(h(e))_i=(k^\star(s_{\mathcal{G}}(e)))_i=k(s_{\mathcal{G}}(e)_i)=k(v)\]
		\item $v$ occurs in $t_{\mathcal{H}}(e)$ for some $e\in E_{\mathcal{H}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{H}}}(t_{\mathcal{G}}(e))$ be such that $v=t_{\mathcal{G}}(e)_i$, then we have
		\[g(v)=g(t_{\mathcal{G}}(e)_i)=(g^\star(t_{\mathcal{G}}(e)))_i=t_{\mathcal{H}}(f(e))_i=t_{\mathcal{H}}(h(e))_i=(k^\star(t_{\mathcal{G}}(e)))_i=k(t_{\mathcal{G}}(e)_i)=k(v)\]
	\end{itemize}
	In both cases $\hat{g}(v)=k(v)$ and so $\hat{g}= k$ as wanted.
\end{rem}

@@

	Suppose that there exists $\mathcal{G}$ and arrows $(p_1, p_2)\colon \mathcal{Q}\to \mathcal{G}$ and $(q_1, q_2)\colon \mathcal{K}\to \mathcal{G}$ which make the square a pushout in $\shyp$. 
	By construction we have
	\[s_{\mathcal{G}}(q_1(e)) = q^\star_2(s_{\mathcal{K}}(e)) = q^\star_2(j^\star_2(v))=p^\star_2(i^\star_2(v)) = p^\star_2(t^\star_{\mathcal{Q}}(e))= t_{\mathcal{G}}(p_2(e))\]
	and so $\mathcal{G}$ cannot be in $\shyp$, yielding a contradiction.
\end{exa}
\end{minipage} \hfill \begin{minipage}[r]{.45\linewidth} \vspace{-2cm}
\xymatrix@R=6pt@C=25pt{\begin{tikzpicture}
		\node[style=node](v) at (0, 0){};
	\node[style=none] at(0, 0.3){$v$};
	\end{tikzpicture}\ar[d]_{(i_1, i_2)} \ar[r]^-{(j_1, j_2)}  &
	\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$j_2(v)$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$w$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, 0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw[style=none] (v) to (in.east); 
		\draw[style=none] (out.east) to (u.west);
	\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[d]^{(q_1, q_2)} \\\begin{tikzpicture}[baseline=(v.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v) at (-1, 0){};
			\node[style=none] at(-1, 0.3){$u$};
			\node[style=node](u) at (1, 0){};
			\node[style=none] at(1, 0.3){$i_2(v)$};
			\node[style=small box] (edge) at (0, 0) {};
			\node[style=none] (edge) at (0, -0.5) {$e$};
			\node[style=none] (in) at (-0.3, 0) {};
			\node[style=none] (out) at (0.2, 0){};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw[style=none] (v) to (in.east); 
			\draw[style=none] (out.east) to (u.west);
		\end{pgfonlayer}
	\end{tikzpicture} 
	\ar@{.>}[r]_-{(p_1, p_2)}&
	\mathcal{G}
	}
\end{minipage}

If we restrict to pruned separated hypergraph the previous issue disappeares.  The crucial observation is that the a pruned separated hypergraph is characterized by its set of hyperedges and the length of their sources and targets. See \Cref{proof:pruned} for a proof.

\begin{restatable}{lem}{pruned}\label{lem:pruned}
$\pshyp$ is equivalent to $\Set/(\mathbb{N}\times \mathbb{N})$.
\end{restatable}
\begin{cor}\label{cor:pruned}
	$\pshyp$ is adhesive.
\end{cor}

@@

\begin{thm} Let $\mathcal{G}$ be an hypergraph, then %the following fact hold
%	\begin{enumerate}
%		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		%\item 
		$\lmo_{\mathcal{G}}$ is quasiadhesive.
%	\end{enumerate}
\end{thm}

@@

\subsubsection{title}
\paragraph{title}
\paragraph{title}
%\input{Hypergraphs_with_equivalences}
%\input{Labeled_hyp_with_eq}

@@

\subsubsection{title}

@@

	Now let $(h,k)$ be another morphism $\mathcal{H}\to \sp(\mathcal{G})$ such that $(\id{E_{\mathcal{G}}}, \epsilon_{G})\circ P(h, k)=(f,g)$. This entails immediately that $h =f$ and so uniqueness follows from \Cref{rem:mprun}.	
\end{proof}

@@

\pruned*

\begin{proof}{\label{proof:pruned}}
	Let $\mathcal{G}$ be a pruned and separated hypergraph. We can define $F(\mathcal{G})$ as $(\lgh_{V_{\mathcal{G}}}\circ s_{\mathcal{G}}, \lgh_{V_{\mathcal{G}}}\circ t_{\mathcal{G}}) \colon E_{\mathcal{G}}\to \mathbb{N}\times \mathbb{N}$. Given a morphism $(h,k)\colon \mathcal{G}\to \mathcal{H}$ and using \Cref{rem:le} we	have
	\[(\lgh_{V_{\mathcal{H}}}\circ s_{\mathcal{H}}, \lgh_{V_{\mathcal{H}}}\circ t_{\mathcal{H}})\circ h = (\lgh_{V_{\mathcal{H}}}\circ s_{\mathcal{H}}\circ h, \lgh_{V_{\mathcal{H}}}\circ t_{\mathcal{H}}\circ h)=(\lgh_{V_{\mathcal{H}}}\circ k^\star\circ s_{\mathcal{G}}, \lgh_{V_{\mathcal{H}}}\circ k^\star\circ t_{\mathcal{G}})= (\lgh_{V_{\mathcal{G}}}\circ s_{\mathcal{G}}, \lgh_{V_{\mathcal{G}}}\circ t_{\mathcal{G}})\]
	Thus we can define $F(h,k)$ simply as $h$ to get a functor $\pshyp\to \Set/(\mathbb{N}\times \mathbb{N})$. Let us show that it is an equivalence.
	\begin{itemize}
		\item $F$ is faithful. This follows at once from \Cref{rem:prun}.
		\item $F$ is full. Let $k\colon E_{\mathcal{G}}\to E_{\mathcal{H}}$ be an arrow such that 
		\[(\lgh_{V_{\mathcal{H}}}\circ s_{\mathcal{H}}, \lgh_{V_{\mathcal{H}}}\circ t_{\mathcal{H}})\circ f =(\lgh_{V_{\mathcal{G}}}\circ s_{\mathcal{G}}, \lgh_{V_{\mathcal{G}}}\circ t_{\mathcal{G}})\]

		Now, for every $e \in E_\mathcal{G}$, let $\iota^s_e\colon V^s_e\mto V_{\mathcal{G}}$ and $\iota^t_e\colon V^t_e\mto V_{\mathcal{G}}$ be the inclusions of
		\[V^s_e=\{v \in V_{\mathcal{G}} \mid \ind{G}(v,e)=1\} \qquad V^t_e=\{v \in V_{\mathcal{G}} \mid \outd{G}(v,e)=1\}\]

		\noindent
		\begin{minipage}[l]{.78\linewidth}
		Since $\mathcal{G}$ is pruned and separated, $(V_{\mathcal{G}}, \{\iota^s_e\}_{e\in E_{\mathcal{G}}}+\{\iota^t_e\}_{e\in E_{\mathcal{G}}})$ is a coproduct. Given $v\in V^s_e$ and $w\in V^t_e$, let $i_v$ and $k_w$ be the unique indexes such that $\iota^s_e(v)=s_{\mathcal{G}}(e)_{i_v}$ and $\iota^t_e(w)= t_{\mathcal{G}}(e)_{k_w}$ and define:
			\[\begin{split}
				g^s_e\colon V^s_{e} &\to V_{\mathcal{H}}\\
				v&\mapsto  s_{\mathcal{H}}(f(e))_{i_v}
			\end{split} \qquad \begin{split}
				g^t_e\colon V^t_{e} &\to V_{\mathcal{H}}\\
				w&\mapsto  t_{\mathcal{H}}(f(e))_{k_w}
			\end{split}\]
			so that there is an arrow $g\colon V_ {\mathcal{G}}\to \mathcal{H}$ fitting in the diagram aside.
		\end{minipage}\hfill
		\begin{minipage}[r]{.2\linewidth}
		\xymatrix@R=14pt@C=14pt{&V^s_{\mathcal{G}} \ar[dr]^{g^s_e} \ar@{>->}[dl]_{\iota^s_e}\\ V_{\mathcal{G}} \ar@{.>}[rr]^-{g} && V_{\mathcal{H}}}		
		\xymatrix@R=14pt@C=14pt{&V^t_{\mathcal{G}} \ar[dr]^{g^t_e} \ar@{>->}[dl]_{\iota^t_e}\\ V_{\mathcal{G}} \ar@{.>}[rr]^-{g} && V_{\mathcal{H}}}		
		\end{minipage}

		Given $e\in E_\mathcal{G}$, for every $1\leq i \leq \lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $1\leq k \leq \lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$ we can compute to get:
		\begin{align*}
		(g^\star(s_{\mathcal{G}}(e)))_i&=(g^\star(s_{\mathcal{G}}(e)))_i=g(s_{\mathcal{G}}(e)_i)=s_{\mathcal{H}}(f(e))_i\\
		(g^\star(t_{\mathcal{G}}(e)))_k&=(g^\star(t_{\mathcal{G}}(e)))_k=g(t_{\mathcal{G}}(e)_k)=t_{\mathcal{H}}(f(e))_k
		\end{align*}

		Hence $(f,g)$ is a morphism $\mathcal{G}\to \mathcal{H}$ such that $F(f,g)=f$.
		\item $F$ is essentially surjective. Consider a function $(f_1, f_2)\colon X\to \mathbb{N}\times \mathbb{N}$. Given $x\in X$,  we can consider $f_1(x)$ and $f_2(x)$ as ordinal sets and take the coproduct $(V, \{\iota^1_x\}_{x\in X}+ \{\iota^2_x\}_{x\in X})$ of the family $\{f_1(x)\}_{x\in X}+\{f_2(x)\}_{x\in X}$. Define:
		\[\begin{split}  
			s\colon X&\to V^\star\\
			x &\mapsto \prod_{n=0}^{f_1(x)-1}\iota^1_x(n)
		\end{split} \qquad  \begin{split}  
		t\colon X&\to V^\star\\
		x &\mapsto \prod_{n=0}^{f_2(x)-1}\iota^2_x(n)
		\end{split}\]
		Let $\mathcal{G}$ be $(X, V, s, t)$, by construction $F(\mathcal{G})=(f_1, f_2)$ and we can conclude.
		 \qedhere 
	\end{itemize}

@@

\section{Some properties of comma categories}
In this section we briefly recall the definition of the comma category \cite{mac2013categories} and some of its properties.
\begin{defi}\index{category!comma -}
	Let $L\colon \A\to \X$ and  $R\colon \B\rightarrow \X$ be two functors with the same codomain, the \emph{comma category} $\comma{L}{R}$ is the category in which:

\noindent
\begin{minipage}[l]{.8\linewidth}
	\begin{itemize}
		\item objects are triples $(A, B, f)$ with $A\in \A$, $B\in \B$, and $f\colon L(A)\rightarrow R(B)$; 
		\item a morphism $(A, B, f)\rightarrow (A', B', g)$ is a pair $(h, k)$ with $h\colon A\rightarrow A'$ in $\A$ and $k\colon B\rightarrow B'$ in $\B$ such that the diagram aside commutes.\end{itemize}
\end{minipage}\hfill
\begin{minipage}[r]{.25\linewidth}
	\xymatrix@R=15pt{L(A) \ar[r]_{L(h)} \ar[d]_{f} & L(A') \ar[d]^{g}\\ R(B) \ar[r]^{R(k)}& R(B')}
\end{minipage} 
\end{defi} 
We have two forgetful functors 	$U_L\colon \comma{L}{R}\to \A$ and $U_R\colon \comma{L}{R} \to \B$ given respectively by
\[
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{A}{h}{A'}
\end{split}\quad 
\begin{split}
	\functor[l]{(A,B, f)}{(h,k)}{(A', B', g)}
	& \functormapsto
	\rfunctor{B}{k}{B'}
\end{split}
\]

	Given $L\colon \A\to \X$ and $R\colon \B\to \X$, we can also consider their duals $L^{op}\colon \A^{op}\to \X^{op}$ and $R^{op}\colon \B^{op}\to \X^{op}$.  An arrow $f\colon L(A)\to R(B)$ in $\X$ is the same thing as an arrow $f\colon R^{op}(B)\to L^{op}(A)$ in $\X^{op}$. Thus $\comma{L}{R}$ and $\comma{R^{op}}{L^{op}}$ have the same objects. Moreover, the left square below commutes in $\X$ if and only if the right one commutes in $\X^{op}$.
\[\xymatrix@R=15pt{L(A) \ar[r]^{L(h)} \ar[d]_{f} & L(A') \ar[d]^{g} & R(B') \ar[d]_{g} \ar[r]^{R(k)}& R(B) \ar[d]^{f}\\ R(B) \ar[r]_{R(k)}& R(B') & L(A') \ar[r]_{L(h)}  & L(A)}\]

@@

	\begin{minipage}[l]{.75\linewidth}
	Thus there exists $f\colon L(A)\rightarrow R(B)$ fitting in the diagram on the right. Notice that $f$ is the unique arrow in $\X$ which makes $\left(a_D, b_D\right)$ an arrow $\left(A_D, B_D, f_D\right)\to \left(A, B, f\right)$ of $\comma{L}{R}$. If we show that $\left((A, B, f), \left\{(a_D, b_D)\right\}_{D\in \D}\right)$ is colimiting for $F$ we are done.
	\end{minipage}
	\hfill
	\begin{minipage}[r]{.2\linewidth}
	\xymatrix@R=10pt{ L(A_D) \ar[r]^{L(a_D)} \ar[d]_{f_D}& L(A) \ar[d]^{f}\\ R(B_D) \ar[r]_{R(b_D)} &R(B)}
	\end{minipage}

@@

We can use \Cref{lim} to characterize monos in comma categories. 
\begin{cor}\label[corollary]{cor:mono}
	If $R$ preserves pullbacks then an arrow $(h,k)$ in $\comma{L}{R}$ is mono if and only if both $h$ and $k$ are monos.
\end{cor}
\begin{proof}
	$(\Rightarrow)$  If $(h,k)\colon (A,B,f)\to (A', B', g)$ is a mono then the first rectangle below is a pullback in $\comma{L}{R}$. By \Cref{lim} then also the other two squares are pullbacks, respectively in $\A$ and $\B$, proving that both $h$ and $k$ are monos.
	\[\xymatrix@R=15pt{(A, B, f)  \ar[rr]_{\id{(A,B,f)}} \ar[d]_{\id{(A,B,f)}}&& (A, B, f)  \ar[d]^{(h,k)}& A \ar[r]_{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]_{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ (A, B, f)  \ar[rr]^{(h,k)}&& (A', B', g) &A \ar[r]^{h} & A' & B \ar[r]^{k} & B'}\]
	\smallskip\noindent
	\begin{minipage}[l]{.3\linewidth}
		 \xymatrix@R=13pt{A \ar[r]_{\id{A}} \ar[d]_{\id{A}} & A \ar[d]^{h} & B \ar[r]_{\id{B}} \ar[d]_{\id{B}} & B  \ar[d]^{k}\\ A \ar[r]^{h} & A' & B \ar[r]^{k} & B'}
	\end{minipage} 
	\hfill
	\begin{minipage}[r]{.65\linewidth}
		$(\Leftarrow)$ Since $h$ and $k$ are monos then we have the two pullback squares on the left. Thus by \Cref{lim} the pullback of $(h,k)$ along itself has isomorphisms as projections and so $(h,k)$ is mono. \qedhere
	\end{minipage}
\end{proof}

@@

\begin{prop}\label[proposition]{prop:left}
	If $\A$ has initial objects and $L$ preserves them, then the forgetful functor $U_R\colon \comma{L}{R}\to \B$ has a left adjoint $\Delta$.
\end{prop}
\begin{proof} For an object $B\in \B$ we can define $\Delta(B)$ as $(0, B, ?_{R(B)})$, for $0$ is an initial object in $\A$ and $?_{R(B)}$ is the unique arrow $L(0)\to R(B)$. Let $\id{B}\colon B\to U_R(\Delta(B))$ be the identity, and suppose that a $k\colon B\to U_R(A, B', f)$ in $\B$ is given.

	\noindent
\begin{minipage}[l]{.2\linewidth}
	\xymatrix@C=30pt@R=8pt{L(0) \ar[r]_{L(?_A)} \ar[d]_{?_{R(B)}} & L(A) \ar[d]^{f}\\ R(B) \ar[r]^{R(k)}& R(B')}
\end{minipage}  \hfill 
	\begin{minipage}[r]{.75\linewidth}
		By initiality of $0$, there is only one arrow $?_A\colon 0\to A $ in $\A$ and, 
		since $L$ preserves initial objects, the square aside commutes. Thus $(?_A,k)$ is the unique morphism $\Delta(B)\to (A, B', f)$ such that $U_R(?_A,k)=k$. \qedhere 
	\end{minipage}

\end{proof}
Dualising we get immediately the following.
\begin{cor}If $\B$ has terminal objects preserved by $R$, then $U_L\colon \comma{L}{R}\to \A$ has a right adjoint.
\end{cor}
